Predictive modeling using genomic annotations

### 2018-09-10 Manuscript/manuscript.Rmd
updates

The advent of various genome-wide technologies, such as high-throughput conformation capture (Hi-C), have revealed how the spatial organization of the human genome may affect several epigenetic functions (Aiden et al.). Analyses have shown that the genome is tightly compacted into distinct compartments.  There exist regions within these compartments that are highly conserved and self-interacting, and termed topologically associating domains (TADs) (Dixon et al.). Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD (Symmons et al.). This suggests that the boundaries of TADs may play a role in restricting the function of elements such as enhancers, thereby impacting the transcription of genes.

More recently, it has been discovered that insulators have a primary role in orchestrating the topological arrangement of higher-order chromatin architecture (Phillips-Cremins et al). Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long range chromatin contacts (Zuin et al.). Likewise, it was found that DNase I-hypersensitive sites were enriched at the boundaries of domains and correlated with the CTCF signals (Hong et al) [??? They identify more]. On the other hand, it is unclear how other types of regulatory elements such as histone modifications, which are associated with transcriptional initiation and open chromatin structure, are related to TAD boundaries.

The distinct patterns of some of these different proteins and functional elements point toward the opportunity of computational approaches in predicting the development of TAD boundaries. However, due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role of specific sets of these features on the folding of chromosomes. Furthermore, many widely used methods ignore key characteristics of the data that may hinder the performance of certain parametric models [??? Like what? It may be best to devote a paragraph clearly describing them. Then, describe the Mourad and other approaches]. One such group have proposed a multiple logistic (MLR) model used to identify the most influential proteins with regards to TAD boundaries (Mourad et al.). They also provide an MLR model with LASSO estimated coefficients that they believe is better suited for predictors that may be correlated. However, key aspects of the data were ignored with each of these models, such as the sparsity of domain borders throughout the genome. Likewise, data pre-processing techniques such as normalization, the elimination of low variance predictors, and variable selection techniques were not considered. Furthermore, due to the large number of genomic features that can be considered, accounting for the relationship among features with interaction effects becomes computationally infeasible.

Data pre-processing techniques such as normalization and the elimination of low variance predictors are an integral part of model fitting with regards to genomic data [??? Clarity. Define 1) the problem, and 2) ways to solve it, which is modeling. Then, for modeling, describe problems connecting genomics+machine learning terms]. Continuous predictors tend to be highly shewed. Therefore, a common practice is to perform a log base 2 transformation. It is also sometimes necessary to standardize continuous predictors, especially if they are on different scales. It is unclear, however, which combination of these two concepts improves prediction. Additionally, binary predictors may only be concordant with the outcome of interest in very rare cases. This contributes to predictors with near-zero variances. Eliminating these predictors prior to model fitting can reduce both the noise and the computational speed of the model. We use the nearZeroVar function provided in the caret package to determine which predictors have near-zero variance [??? This sentence is Methods].

TADs can be up to a million base pairs in length and therefore the boundaries are sparse throughout the genome. Sparse data can create heavily imbalanced classes and may affect prediction performance of classification algorithms. There are many techniques that can be used to more evenly balance such data. These include oversampling the minority class, under sampling the majority class, and some combination of both. One such technique is referred to as SMOTE, which stands for Synthetic Minority Over-sampling TEchnique. SMOTE is a function in the DmWR package in R and incorporates both under-sampling and over-sampling. It has been shown that SMOTE out performs simple under-sampling of the majority class for some machine learning algorithms (Nitesh et al). We instead propose a method of taking multiple bootstrap under-samples from the majority class and then aggregating performance metrics by taking the average across all iterations. We then compare this method to SMOTE.

Modeling genomic data often involves a large number of predictors as well. As a result, different machine learning models can benefit from variable selection techniques to reduce the feature space, and thereby improve computational speed. There are several known selection techniques in the field of machine learning including forward, backward, and stepwise selection. These are known as wrapper methods because they measure the usefulness of a subset of features by actually training a model on it. Additionally, we incorporate recursive feature elimination as a method of variable reduction and compare it to the rest.

Furthermore, the model proposed by Mourad et al. falls short of addressing issues such as cross-validation and variable importance, which can be handled more efficiently through ensemble models like random forests and gradient boosting machines.  Therefore, in addition to proposing a novel pipeline that addresses irregularities in the data mentioned above, we also apply a random forest model in order to find the key molecular drivers most associated with TAD boundaries. A random forest was chosen because of its ability to handle potentially correlated variables as well as its inability to overfit. We then compare the performance of the random forest model to the models proposed by Mourad et al.

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect several epigenetic functions (Aiden et al.). Analyses have shown that the genome is tightly compacted into distinct compartments.  There exist regions within these compartments that that have been shown to be highly conserved and self-interacting, and termed topologically associating domains (TADs) (Dixon et al.). Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD (Symmons et al.). This suggests that the boundaries of TADs may play a role in restricting the function of elements such as enhancers, thereby impacting the transcription of genes.

More recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangement of higher-order chromatin architecture (Phillips-Cremins et al). Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long range chromatin contacts (Zuin et al.). Furthermore, other regulatory elements like DNase I hypersensitive sites, the H3K36 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions (Hong et al). Each of these associated with more open chromatin which alter the accessibility of genes for transcription.   [??? They identify more; !!! included mention of histone mods. and TFs]. The distinct patterns of some of these different proteins and functional elements point toward the opportunity of computational approaches in predicting the development of TAD boundaries.

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al). Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. It is unclear if distances or overlaps between genomic bins and regions of functional elements contain the most information. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation, and it is important that the proposed model is able to handle multiple correlated features. [??? Like what? It may be best to devote a paragraph clearly describing them. Then, describe the Mourad and other approaches; !!! stated the issues, now will discuss current methods].

In a previous study (Mourad et al) both a multiple logistic regression model and a LASSO regularized regression model were applied to a wide array of architectural proteins from the GM12878 cell line to predict TAD boundaries. The set of features consisted of whether or not the coordinates of a functional element overlaped with a particular genomic bin (yes or no). In the case of a partial overlap, a percentage of overlap was used. However, the class imbalance issue was not addressed. Additionally, neither model is suited to handle multiple correlated predictors. Furthermore, ROC curves and AUCs were used to evaluate model performance. This is problematic, especially for highly imbalaned classes because it dismisses how well the model is able to identify true positive cases. In another study (Huang et al), a more robust model in the form of a Bayesian Additive Regression Trees (BART) model was used. The set of features consisted of a score that averaged the number of sequence reads for each functional element over a particular bin for the GM12878 cell line. Here, the balanced classes were manually created by dividing the positive and negative classes into two equal subsets used for model training and testing. However, it is unclear what balancing technique was used. In addition to this, the model was only performed on a small set of histone marks while ignoring wider classes of epigenomic elements. Lastly, in a more recent study (Hong et al), a bayesian rigde regression model was iteratively used on a set of transcription factor binding sites, transcription start sites, and DNase I hypersensitive sites from the hESC cell line until the best combination of features led to the optimal prediction of TAD boundaries.  Here, the features of the model consisted of the number of occurences of each genomic element within each genomic bin. Balanced classes were manually created by randomly sampling from the set of bins that did not contain a TAD boundary. Given the structure of the search algorithm proposed, though, it can be computationally inefficient to have to iterate through a large set of genomic features to find the particular set that contributes to the best prediction. Because of that, only a maximum set of 10 genomic features were considered at any given time.

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Additionally, a notable omission was the use of distances between regions of the functional elements and genomic bins. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in addition to proposing a novel data pre-processing pipeline that addresses irregularities in the data, we also apply an elastic-net model (and/or random forest?) using both distances and counts of overlaps between bins and features as predictors in our model. We aim to assess both the predictive performance of the model and the influence of genomic features on the borders that demarcate topological domains.[??? Clarity. Define 1) the problem, and 2) ways to solve it, which is modeling. Then, for modeling, describe problems connecting genomics+machine learning terms; !!!identified problem and aim]
